Details of specific functions

### 2018-01-04 paper/response_to_reviewers.tex

\item{\textcolor{blue}{Another concern I have is about how this R package sits within the current analytical toolkit available.
How does this new package relate to existing packages available for R?
Claddis and geomorph are mentioned and discussed because they deal with bespoke data formats that need to be dealt with prior to analysis.
But are there others?
How does the package relate to the ‘vegan’ R package for multivariate analysis of ecological communities?
Within there are metrics for community dissimilarity, metrics from ordinations etc.
Vegan is a popular package for analysis of multivariate datasets and seems to be relevant here with similar functionalities but not discussed.
Similarly, the R package ‘diverse’ has a function called disparity, and the ‘geiger’ function disparity implements the disparity-through-time. }}
\label{add_geiger_vegan}

I have added two wrapper functions to use the popular functions \texttt{vegan::adonis} and \texttt{geiger::dtt} to the package (\texttt{dispRity.adonis} and \texttt{dispRity.dtt}).
Also, I have generalised the distances calculations using the \texttt{vegan::vegdist} function that allows for more distances than \texttt{stats::dist} to be calculated.

I've updated the manual and the example to describe these new functions:

\textit{Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA (PERMANOVA; Anderson, 2001b) through the adonis.disprity function that is a wrapper of the vegan::adonis function (Oksanen et al., 2007) for dispRity objects [TABLE 4]. To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset. Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test. Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 359-369.

@@

I have however modified the available models to better describe this limitation and I have added two new models that would approximate ancestral reconstruction along edges.
I have modified the name of the two previous ``punctuated'' and ``gradual'' models to ``random'' and ``proximity'' to reflect more accurately how they select either the ACCTRAN or DELTRAN information.
I've added two more accurate models (now named ``equal.split'' and ``gradual.split'') to reflect the changes along the branches.
These two algorithm now allows to use both the ACCTRAN and DELTRAN information with a linked probability to use either of them per bootstrap draw.
For example, the ``equal.split'' model will have a 0.5 probability of using either node information for each bootstrap draw; the ``gradual.split'' model transforms the branch length were the slice occurs into a probability of drawing either of the nodes information (e.g. if the slices occurs 3/4 along the branch from A to B, the model will draw the information from A and B with a probability of respectively 0.25 and 0.75).
For example, the median disparity over 1000 bootstrap pseudo-replication along the edge (AB) will integrate 0.75 information from A and 0.25 from B thus approximating the ``intermediate'' between A and B (when averaged ).

I've added these models descriptions in the manual and in the manuscript:

\begin{itemize}
    \item \textit{``random'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor}
    \item \textit{``proximity'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length}
    \item \textit{``equal.split'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
    \item \textit{``gradual.split'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
\end{itemize}

\textit{Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} lines 169-183.

@@

I've changed ``subsample'' to ``subset'' throughout the package, the manual and the paper.

I also agree with this comment: this method is a trade-off between the number of slices.
(1) too few slices decrease resolution and veer towards the time binning (which could increase "significance" by comparing too different time subsamples);
(2) too many slices will increase the resolution be will create much more pseudo replication and thus increase type I error.

I've added a warning in the manual and the manuscript:

\textit{Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).} line 180-183.

@@

\textit{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements (e.g. the multidimensional space can be a distance matrixl; Close et al., 2015).} lines 109-111.

\item{\textcolor{blue}{Picking up on the first point from above, there are also additional post-ordination disparity metrics that have been used in the literature but are not available here. For example, hypercube volume, minimum spanning tree length etc. I would also urge the author to cite the work of Mike Foote, who made some early important contributions to disparity in general, and has written papers that discuss some of these potential metrics (e.g., Foote 1997, Annual Review of Ecology and Systematics, 28, 129–). I believe Wills (2001, cited) also discusses a long list of metrics.}}

I agree, there are many more metrics commonly used for disparity analysis not implemented in this package.
This was a conscious choice: due to the really high number of metrics available (especially when combining different ones), I designed the package to be modular hence allowing users to implement their own metrics (see response to reviewer 1 minor comment \ref{table_missing}).
I have add the hypercube volume description as an example of user defined metric in the manuscript:

\textit{When multiple functions are passed to the} \texttt{metric} \textit{argument, they are sorted by dimension-level and applied in decreasing order to the data.
For example, if the metric is defined as} \texttt{metric = c(prod, ranges)} \textit{(the hypercube volume), the} \texttt{ranges} \textit{function (dimension-level 2) is first applied to data (the range of each dimensions) and the function} \texttt{prod} \textit{is then applied to the results (}\texttt{prod(ranges(data))}\textit{).
One can also directly pass a function description to the} \texttt{metric}\textit{ argument (e.g.} \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} \textit{ for the average squared pairwise distance).} lines 130-135.

@@

Similarly to previous comment, I will also add the permutation test based on the \texttt{PAST} package in a future version of the package.
Furthermore, I have added an illustrative example in the \texttt{plot.dispRity} manual on how to add disparity data (continuous or discrete) to the \texttt{geoscalePlot} and \texttt{geoscaleBox} functions thus allowing users to make nice ready-to-publish figures using the \texttt{geoscale} package.

\item{\textcolor{blue}{and options for the user to only use a limited set of ordination axes (matrix columns), e.g., to avoid using higher axes with negative eigenvalues.}}

This option is implemented in the \texttt{boot.matrix} and \texttt{dispRity} functions allowing users to set a \texttt{dimensions} argument to either a proportion or a value of dimensions to use (e.g. \texttt{0.95} or \texttt{10} to use only the 95\% of the dimensions or the 10 first ones respectively). I've added a line highlighting this feature that is indeed often used:

\textit{Note that this function also allows to work on only a subset of dimensions via the} \texttt{dimensions} \textit{argument (e.g. if only the $n$ first dimensions must be considered).} lines 134-135.

@@

\item{\textcolor{blue}{L116 - "??"!}}

Sincerest apologies! See reviewer 1 minor comment \ref{table_missing}.

\item{\textcolor{blue}{L130 - I don't think the term "possible observed" makes sense. Either the author means all the observed morphologies *or* all the possible morphologies. I suggest the former is safer as I think it likely that possible morphologies introduce sufficient complexity we can never really say this is true. For example, if we alter the number of elements in our sample (removing a species say) the ordination/morphospace we produce will differ, but the range of possible morphologies should not.}}

I agree and have changed it to ``\textit{all the observed morphologies}'' line 150

@@

\item{\textcolor{blue}{L134 - Should probably be "an argument".}}

I have changed the start of the sentence to ``\textit{The first argument is the matrix}'' line 153.

@@

\item{\textcolor{blue}{L136 - It is not immediately obvious to me why a phylogeny is required here. This only becomes clearer later (time-slicing) so maybe a "heads up" parenthetic (i.e., "see below") could be added here.}}

I've modified the sentence in bracket to
``\textit{a dated phylogeny of the elements present in the morphospace - see below}'' line 155.

@@

\label{ANOVA}

I agree that this ANOVA is a garbage-in-garbage-out test and used it more for illustrating the package's possibilities rather than show-casing disparity analysis best (worst?) practices.
Following this reviewer's comments with which I totally agree, I have completely removed the ANOVA example (from the paper, the manuals and the vignette), replacing it in the paper with a PERMANOVA (based on \texttt{vegan::adonis}) which is more conventional.
As this reviewer points out, the ANOVA is indeed performed on the bootstrapped data rather than the observed data which, of course, artificially (and wrongly) increases the statistical power.
Although the ANOVA is not mentioned any more, I have however left the ability for \texttt{test.dispRity} to deal with \texttt{lm} type tests (with a warning being triggered).

I have implemented the reviewers' suggestion similary to the ``percentile bootstrap'' idea, allowing \texttt{test.dispRity} to work on either the pooled bootstrap distribution through the \texttt{concatenate = TRUE} argument (this was indeed erroneous in the previous ANOVA example but could be justifiable in other cases; e.g. when comparing distribution overlap probabilities) or by applying the test for each bootstrap pseudo-replicate with \texttt{concatenate = FALSE}. I've clarified this in the manuscript:

\textit{Additionally, the test is here performed on the pooled bootstrapped pseudo-replications which can increase the type I error.
It is possible to compare each bootstrap in a pairwise way without pooling the data by using the} \texttt{concatenate = FALSE} \textit{argument.
The results will then be a distribution of statistic and p values.} lines 323-327.

@@

I have changed the \texttt{test.dispRity} to now output the name of the statistic and the parameter in the output list (e.g. \texttt{statistic: t} and \texttt{parameter: df} for the example provided by the reviewer).
I have also highlighted in the manuscript one of the features of the function allowing to output the results in a format more familiar to the user using the option \texttt{details = TRUE} (returning the raw test results):

\textit{Note that by default the function only outputs the test's statistic, parameter (if parametric) and the p value. However, the raw test results can also be output using the option} \texttt{details = TRUE} \textit{in the function above.} lines 319-320.

@@

\section{Reviewer 4 (Gavin Simpson)}

\begin{enumerate}

\item{\textcolor{blue}{I have several concerns regarding the content in the manuscript, however, primarily how it is pitched.
It seems unclear, to me at least, who the intended audience is?
At various points I get the impression that the author intends for the package to be useful, and therefore, presents it, to readers beyond those studying morphometry etc.
There are frequent references to ecology and related disciplines where similar approaches are used.
Yet, those in these broader fields will struggle to follow what is presented as the manuscript exclusively uses "morphospaces" and related terminology that will only be relevant to a specific subset of readers.If you want to aim for a more general audience, consider rewriting the manuscript in more general terminology; samples in rows, variables in columns, for example.
Perhaps include a table to relate terminology used in morphometry with that more commonly encountered in community ecology.
Consider adding an ecological example and removing, if length is an issue, content on the manual and vignette --- we don't really need a paragraph on that, just a sentence to highlight it exists.}}

Due to space constraint for this format of MEE manuscript, I have left the references to ecology in the introduction of the manuscript but removed any further references.
However, I have expanded the ecological example in the vignette which now include a test similar to the one in Diaz \textit{et al} 2016 (Science).
I therefore hope that the mention of ecology in the introduction will encourage ecology-focused researcher to explore the manual.

\textit{The following example is based on the discrete morphological data described above. More examples, including based on other types of data are available in the package vignette.} lines 271-273

Additionally, I've also added a Glossary table as suggested and generalised disparity data description as suggest by this reviewer's minor comment \ref{rows}.

@@

\item{\textcolor{blue}{The referencing is poor; why cite Close et al for the concept of pairwise distances?
I'm pretty sure PCO existed before Brusatte et al, and PCA certainly before 2012, etc.
Throughout I was frustrated with this; the referencing ignores a century of work outside this specific field.
I appreciate there merit in citing relevant literature for your field, but that doesn't absolve one of a duty to accurately attribute ideas to those that came up with them.
If you are unfamiliar with developments in ideas beyond your particular field, I would suggest Legendre \& Legendre's *Numerical Ecology* textbook as a good source for original work relevant to what is presented here.}}

I thank this reviewer for sharing the excellent Legendre \& Legendre reference (that was definitely a reference I was missing) and I've accurately credited the statistical methods throughout the manuscript - see my response to reviewer 2, minor suggestion \ref{PCO_cites}.

\item{\textcolor{blue}{The comparison or critique of packages that, in the eyes of the author, restrict users grated a little with me. As a developer of one of the packages named, I feel that particular critique is unwarranted for several reasons;
\begin{itemize}
\item i) **vegan** doesn't restrict you at all, as with anything in R you are free to pick and choose ordination methods, dissimilarity measures etc and then do what you want with the scores,
\item ii) **vegan** doesn't provide disparity measures because it was never intended for such usage, but it is general enough that a user could do whatever they want,
\item and iii) like any software, the **dispaRity** author's software included, you have to provide defaults and **vegan**'s defaults reflect the intended audience of community ecology; that's a feature, as it is of **dispaRity**. As such the argument set up on page 6 is some straw man and unnecessary to motivate your package.
\end{itemize}
}}

I apologise for the unfounded critique and I have toned down this sentence (see reviewer 1, minor suggestion \ref{tone_down}).
Following the points raised by this reviewer, I have added functionalities from other packages (namely \texttt{vegan::adonis}) using defaults more appropriate to disparity analysis in palaeobiology (see reviewer 1 major comment \ref{add_geiger_vegan}).

@@

\item{\textcolor{blue}{You don't really define disparity anywhere, except in very loose terms.
This needs making more explicit; I appreciate that disparity is a broad, nebulous concept, but you need to help the reader that may not be familiar with your subject area.
Either that or as mentioned above, refocus the paper on those readers already versed in this particular type of analysis.}}
\label{nebulous}

I have tried to define disparity with more examples and references to be more familiar to a broader range of readers (See reviewer 1, major suggestion \ref{define_disparity}).

\item{\textcolor{blue}{Something that I did miss, which is at the forefront of my mind as a lot of work has gone into **vegan** in this area, is how you handle negative eigenvalues and their eigenvectors?
Typically, one needs to account for these issues, which readily arise when embedding data in an Euclidean space (via PCO) using a non-metric dissimilarity.
Does **dispaRity** handle such problems, and if so how?
If not, is this because those steps are left up to the user to deal with?
If so, there are some issues that cannot be corrected by the user without strongly biasing the results of the analysis.
This is one area where there is the potential for the ideal of being accommodating of users can cause problems downstream if you are not in control of issues like non-metric dissimilarities.
An example where it needs specific handling is in multivariate dispersions (distances around centroids), but even simple ordinations based on dissimilarities require special handling.}}

The package does not deal with negative eigen values in any special way, this task being left to the user to deal with.
In fact, the package simply takes as input a matrix and only considers rows as elements and columns as dimensions; it ignores whether it is an ordinated matrix or not and this, along with the subsequent tasks is left to the users to deal with appropriately.

I have added a warning about that in the manuscript, in the vignette and the package manual (for the functions that allow the matrix as input).

\textit{The matrix is always considered as the multidimensional space and no correction is applied to it (e.g. if ordinated and with negative eigen values, corrections should be applied prior to using the} \texttt{dispRity} \textit{package.)} lines 111-113

@@

\item{\textcolor{blue}{L252 and throughout the code: insert straight quotes. There is a good chance these curly quotes will get replaced with some non-ASCII character and be impossible to paste into R.}}

I've converted all quotes in quoted code lines as ASCII straight quotes.

\item{\textcolor{blue}{In the example around L236--249, can you justify the metric used here? Is the sum of variances along components of an ordination a valid measure of the dispersion in multivariate space of samples? You don't, or I didn't see, reference to original work that justifies this, and it's not something I have come across despite implementing some approaches for testing multivariate dispersion in the **vegan** package. Doesn't this suffer the same problem as confidence bars in an x-y scatterplot where the bars have been defined with reference to only one axis at a time? The solution is a confidence ellipse, which is often quite different to the interval implied by separate confidence interval bars for each axis.}}

This metric (the sum of each ordination axis variance) is commonly used in palaeobiology (since Foote 1991 and Wills et al. 1994) probably due to a ``citation diffusion effect'' (i.e. future authors replicating convincing methods) and partly due to its pretty good behaviour with palaeontological data (e.g. missing data, Ciampaglio 2001).
I now cite these three references lines 293-294 as a justification.

However, as this reviewer highlights, this metric suffers from several problems such as the fact that it ignores the covariance between the dimensions.
This can be a problem when the disparity metric is used for an exact proxy of the multidimensional volume (which, as this reviewer proposes, would be closer to the ellipsoid volume).
However as a standard and comparable disparity metric, the sum of each dimension's variance is mathematically equal to the mean squared pairwise distance, a metric heavily used through it's implementation in \texttt{geiger} (Harmon \textit{et al} 2008 Bioinformatics).

@@

\item{\textcolor{blue}{L294 replace "following up" with "subsequent"}}

I've changed the sentence following this suggestion.

\item{\textcolor{blue}{L323 This statement seems overly vague. What level of confidence is there in this test? The 95\% intervals overlap throughout, and although you don't show then, one can infer that say the 0.2 probability quantile of subsample 2 overlaps with the 0.975 quantile of sub sample 1.}}

I have removed this statement altogether and now directly applied the test, the reader can now read directly to the results of the test (table 5) and the conclusions lines 365-366.

@@

\item{\textcolor{blue}{L326 Where do the 300 samples come from as implied by the residual degrees of freedom. Table 3 suggests 15 + 9 + 13 = 37 samples in this analysis. I'm clearly missing something here. Also, this is an omnibus test, which if we believe the printed values (see previous re DF) only says there is some effect of `subsamples`. Can this analysis be extended to point estimates for mean variance (again assuming you can justify this sum of individual variances method) and post-hoc comparisons of pairwise `subsample` differences, which would be quite useful?}}

The high value of the samples, residuals and degrees of freedom comes from the test being performed on the pooled bootstrap values rather than the raw observations.
I have now removed any mentions to this dodge anova and replaced it with a more traditional PERMANOVA instead (see reviewer 1 major comment \ref{add_geiger_vegan} and reviewer 3, major comment \ref{ANOVA}).

### 2018-01-04 inst/gitbook/02_getting-started.Rmd

>>>>

> Note that whether the matrix is an ordinated one or of any other type, it is always considered as the multidimensional space to analysis and no corrections are applied to it (e.g. if ordinated and with negative eigen values, no correction is used). Always make sure the input is the right multidimensional space to analyse!

### 2018-01-04 inst/gitbook/03_specific-tutorials.Rmd

---
title: "Details of specific functions"
author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

@@

Before each section, make sure you loaded the Beck and Lee (2014) data (see [example data](#example-data) for more details).

```{r}
## Loading the data
data(BeckLee_mat50) ; data(BeckLee_mat99)
data(BeckLee_tree) ; data(BeckLee_ages)
```

@@

@@

```{r, eval=TRUE}
## Creating the two groups as a list
mammal_groups <- list("crown" = c(16, 19:41, 45:50),
                      "stem" = c(1:15, 17:18, 42:44))

@@

Elements can easily be assigned to different groups if necessary!

```{r, eval=FALSE}
## Creating the three groups as a list
mammal_groups <- list("crown" = c(16, 19:41, 45:50),
                      "stem" = c(1:15, 17:18, 42:44).
                      "all" = c(1:50))
```

@@

Note that the `test.dispRity` algorithm deals with some classical test outputs (`h.test`, `lm` and `numeric` vector) and summarises the test output.
It is, however, possible to get the full detailed output by using the options `details = TRUE`.

@@

It is also possible to apply some more *complex* tests that have their own output classes (like `stats::lm`).

The results can then be analysed as usual using the associated `summary` S3 method:

```{r}
## Performing and linear model applied to the same data
(slice_lm <- test.dispRity(disparity_time_slices, test = lm,
                            comparisons = "all"))

## The output is a regular `lm` output
class(slice_lm)

## This output can be summarised using summary
summary(slice_lm)
```

Of course, due to the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
The only condition is that the test can be applied to at least two distributions.
In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm` for example).

>>>>

---
title: "Details of specific functions"
author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "`r Sys.Date()`"
bibliography: [References.bib, packages.bib]
output:
  html_document: default
  pdf_document: default
---

@@

Before each section, make sure you loaded the @beckancient2014 data (see [example data](#example-data) for more details).

```{r}
## Loading the data
data(BeckLee_mat50)
data(BeckLee_mat99)
data(BeckLee_tree)
data(BeckLee_ages)
```

@@

If you want to generate time subsets based on stratigraphy, the package proposes a useful functions to do it for you: `get.bin.ages` (check out the function's manual in `R`)!

@@

```{r, eval=TRUE}
## Creating the two groups (crown and stems)
mammal_groups <- crown.stem(BeckLee_tree, inc.nodes = FALSE)

@@

Like in this example, you can use the utility function `crown.stem` that allows to automatically separate the crown and stems taxa given a phylogenetic tree.
Also, elements can easily be assigned to different groups if necessary!

```{r, eval=FALSE}
## Creating the three groups as a list
weird_groups <- list("even" = seq(from = 1, to = 49, by = 2),
                      "odd" = seq(from = 2, to = 50, by = 2),
                      "all" = c(1:50))
```

The `custom.subsets` function can also take a phylogeny (as a `phylo` object) as an argument to create groups as clades:

```{r, eval=FALSE}
## Creating groups as clades
custom.subsets(BeckLee_mat50, group = BeckLee_tree)
```

This automatically creates 49 (the number of nodes) groups containing between two and 50 (the number of tips) elements.

@@

Note that the `test.dispRity` algorithm deals with some classical test outputs and summarises the test output.
It is, however, possible to get the full detailed output by using the options `details = TRUE`.

@@

Because of the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
The only condition is that the test can be applied to at least two distributions.
In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `lm`-type functions for example).

### NPMANOVA in `dispRity`

One often useful test to apply to multidimensional data is the permutational multivariate analysis of variance based on distance matrices `vegan::adonis`.
This can be done on `dispRity` objects using the `adonis.dispRity` wrapper function.
Basically, this function takes the exact same arguments as `adonis` and a `dispRity` object for data and performs a PERMANOVA based on the distance matrix of the multidimensional space (unless the multidimensional space was already defined as a distance matrix).
The `adonis.dispRity` function uses the information from the `dispRity` object to generate default formulas:

  * If the object contains customised subsets, it applies the default formula `matrix ~ group` testing the effect of `group` as a predictor on `matrix` (called from the `dispRity` object as `data$matrix` see [`dispRitu` object details](#The-dispRity-object-content))
  * If the object contains time subsets, it applies the default formula `matrix ~ time` testing the effect of `time` as a predictor (were the different levels of `time` are the different time slices/bins)

```{r}
set.seed(1)
## Generating a random character matrix
character_matrix <- sim.morpho(rtree(20), 50, rates = c(rnorm, 1, 0))

## Calculating the distance matrix
distance_matrix <- as.matrix(dist(character_matrix))

## Creating two groups
random_groups <- list("group1" = 1:10, "group2" = 11:20)

## Generating a dispRity object
random_disparity <- custom.subsets(distance_matrix, random_groups)

## Running a default NPMANOVA
adonis.dispRity(random_disparity)
```

Of course, it is possible to pass customised formulas if the disparity object contains more more groups.
In that case the predictors must correspond to the names of the groups explained data must be set as `matrix`:

```{r}
## Creating two groups with two states each
groups <- as.data.frame(matrix(data = c(rep(1,10), rep(2,10), rep(c(1,2), 10)),
         nrow = 20, ncol = 2, dimnames = list(paste0("t", 1:20), c("g1", "g2"))))

## Creating the dispRity object
multi_groups <- custom.subsets(distance_matrix, groups)

## Running the NPMANOVA
adonis.dispRity(multi_groups, matrix ~ g1 + g2)
```

Finally, it is possible to use objects generated by `time.subsets`.
In this case, `adonis.dispRity` will applied the `matrix ~ time` formula by default:

```{r}
## Creating time series
time_subsets <- time.subsets(BeckLee_mat50, BeckLee_tree, 
     method = "discrete", inc.nodes = FALSE, time = c(100, 85, 65, 0),
     FADLAD = BeckLee_ages)

## Running the NPMANOVA with time as a predictor
adonis.dispRity(time_subsets)
```

Note that the function warns you that the input data was transformed into a distance matrix.
This is reflected in the Call part of the output (`formula = dist(matrix) ~ time`).

To use each time subset as a separate predictor, you can use the `matrix ~ time.subsets` formula; this is equivalent to `matrix ~ first_time_subset + second_time_subset + ...`:

```{r}
## Running the NPMANOVA with each time bin as a predictor
adonis.dispRity(time_subsets, matrix ~ time.subsets)
```

### `geiger::dtt` model fitting in `dispRity`

The `dtt` function from the `geiger` package is also often used to compare a trait's disparity observed in living taxa to the disparity of a simulated trait based on a given phylogeny.
The `dispRity` package proposes a wrapper function for `geiger::dtt`, `dtt.dispRity` that allows the use of any disparity metric.
Unfortunately, this implementation is slower that `geiger::dtt` (so if you're using the metrics implemented in `geiger` prefer the original version) and, as the original function, is limited to ultrametric trees (only living taxa!)...

```{r}
require(geiger)
geiger_data <- get(data(geospiza))

## Calculate the disparity of the dataset using the sum of variance
dispRity_dtt <- dtt.dispRity(data = geiger_data$dat, metric = c(sum, variances),
                             tree = geiger_data$phy, nsim = 100)

## Plotting the results
plot(dispRity_dtt, fig.width=8, fig.height=8)
```

Note that, like in the original `dtt` function, it is possible to change the evolutionary model (see `?geiger::sim.char` documentation).

### null morphospace testing with `null.test`

This test is equivalent to the test performed in @diaz2016global.
It compares the disparity measured in the observed space to the disparity measured in a set of simulated spaces.
These simulated spaces can be built with based on the hypothesis assumptions: for example, we can test whether our space is normal.

```{r}
set.seed(123)
## A "normal" multidimensional space with 50 dimensions and 10 elements
normal_space <- matrix(rnorm(1000), ncol = 50)

## Calculating the disparity as the average pairwise distances
obs_disparity <- dispRity(normal_space, metric = c(mean, pairwise.dist))

## Testing against 100 randomly generated normal spaces
(results <- null.test(obs_disparity, replicates = 100, null.distrib = rnorm))
```

Here the results show that disparity measured in our observed space is not significantly different than the one measured in a normal space.
We can then propose that our observed space is normal!

These results have an attributed `dispRity` and `randtest` class and can be plotted as `randtest` objects using the `dispRity` S3 `plot` method:

```{r, fig.width=8, fig.height=8}
## Plotting the results
plot(results, main = "Is this space normal?")
```

For more details on generating spaces see the [`space.maker`](#Simulating-multidimensional-spaces) function tutorial.
